---
title: 'Tutorial: Network Structure, Week 2'
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
library(igraph)
library(sna)
library(network)
load("Week2.rdata")
tutorial_options(exercise.timelimit = 10)
```

## Introduction

This tutorial will demonstrate several R commands used to describe aspects of network structure introduced in the Week 2 Lecture:

* Dyads and Dyad Census
* Triads and Triad Census
* Network Transitivity and Clustering
* Path Length & Geodesic
* Component Structure and Membership

Datasets used in this tutorial are the abbreviated Game of Thrones Book *(gotbook)* and Swiss Climate Influence *(climate)* datasets described in the data index for the course.

- *climate.ig* - igraph network object
- *climate.stat* - statnet network object

The *gotbook* network was created from data pulled from the 5 published books, using edges (or ties) between nodes (characters) who are mentioned in the text within close proximity of each other. To retain only ties that are less likely to be incidental, and are likelier to represent a close relationship, ties between characters who appear near each other in the text fewer than 5 times were removed from the network object.

- *gotbook.ig* - igraph network object
- *gotbook.stat* - statnet network object

## Describe the Network Data

First, let's review methods covered in the Week 1 tutorial to describe both datasets.

1. **List and inspect:** List the objects to make sure the network objects are loaded in memory:

```{r ls, exercise=TRUE}
#List the network objects currently available for use

```

```{r ls-solution}
#List the network objects currently available for use
ls()
```

2. **Network Size:** What is the size of the network? Use the commands from week 1 to inspect the count of vertices and edges for each of the igraph objects, or print the network size for the statnet objects. Do this for both the `climate` and `gotbook` objects:

```{r size, exercise=TRUE}
#Find network size (vertex and edge count)

```

```{r size-solution}
#Find network size (vertex and edge count): igraph
vcount(climate.ig)
ecount(climate.ig)

vcount(gotbook.ig)
ecount(gotbook.ig)

#Find network size (vertex and edge count): statnet
print(climate.stat)
print(gotbook.stat)
# Or:
network.size(climate.stat)
network.edgecount(climate.stat)
network.size(gotbook.stat)
# XXX uh-on. This is returning 0, unless na.omit = F is specified.
# The na attribute was removed. Why?
network.edgecount(gotbook.stat)

```

3. **Network features**: Are these networks weighted, directed, and bipartite?

```{r features, exercise=TRUE}
#Find network features (mode, direction, weights)
```

```{r features-solution}
#Find network features: igraph
is_bipartite(climate.ig)
is_directed(climate.ig)
is_weighted(climate.ig)

is_bipartite(gotbook.ig)
is_directed(gotbook.ig)
is_weighted(gotbook.ig)

#Alternatively, find network features: statnet
print(climate.stat)
print(gotbook.stat)
# Or:
is.bipartite(climate.stat)
is.directed(climate.stat)
list.edge.attributes(climate.stat)
```

4. **Network Attributes**: Now, using commands from either statnet or igraph, list the vertex and edge attributes of the `climate` and `gotbook` network data. Be sure to use the correct network data format (and name) for the command you choose to use.

```{r attributes, exercise=TRUE}
#List network attributes (vertex and edge)
```

```{r attributes-solution}
#List network attributes: igraph
igraph::vertex_attr_names(climate.ig)
igraph::edge_attr_names(climate.ig)
igraph::vertex_attr_names(gotbook.ig)
igraph::edge_attr_names(gotbook.ig)

#List network attributes: statnet
network::list.vertex.attributes(climate.stat)
network::list.edge.attributes(climate.stat)
network::list.vertex.attributes(gotbook.stat)
network::list.edge.attributes(gotbook.stat)
```

## Dyad Census
Let's try a full **dyad census**. This gives us the number of dyads where the relationship is:

* Reciprocal (mutual), or `mut`
* Asymmetric (non-mutual), or `asym`, and
* Absent, or `null`

In an undirected graph, all edges are mutual and none are asymmetrical - but this distinction exists when ties are directed.

(Note: older versions of the igraph library gave warnings for undirected graphs, but [this is no longer the case](https://github.com/igraph/igraph/blob/master/CHANGELOG.md#093---2021-05-05). As of July 2022, the R documentation for `igraph::dyad_census` is out of date on this point.)

```{r , echo=TRUE}
#Classify all dyads in the network: igraph
igraph::dyad.census(gotbook.ig)

#Classify all dyads in the network: statnet
sna::dyad.census(gotbook.stat)
```

Note that both igraph and statnet provide functions called `dyad.census`; it is important to understand that this is a coincidence (although an unsurprising one), and that it means that if both libraries are loaded, `dyad.census()` is ambiguous, and will result in the function from whichever library was loaded last in sequence being called. To avoid this - as in any context in R where libraries provide identically-named functions that conflict with each other -  prefix the function name with the specification of the `igraph::` or `sna::` namespace, as appropriate for the object type.

### Exercise: Dyad Census

Now, using commands from either statnet or igraph, conduct a dyad census of the "climate" network. Be sure to use the correct network data format (and name) for the command you choose to use.

```{r dyad, exercise=TRUE}
#Classify all dyads in the network

```

```{r dyad-solution}
#Classify all dyads in the network: igraph
igraph::dyad.census(climate.ig)

#Classify all dyads in the network: statnet
sna::dyad.census(climate.stat)
```

<!-- ##Reciprocity -->

<!-- grecip -->

<!-- edgewise is (m*2)/(m*2+a) -->
<!-- dyadic is (m*2)/(m+a) -->
<!-- dyadic-nonnull is (m)/(m+a) -->

## Triad Census

While it is easy to think of social networks as comprised of dyadic interactions, network scholars are more likely to describe the triad as the true building block of network structure. Dyadic interactions are frequently conducted within the context of one or more third parties that can directly or indirectly affect the outcome. The triad census provides a fundamental descriptive insight into the types of triads found in a particular dataset - refer to the lecture for more details on interpretation.

```{r triad-example, echo=TRUE}
#Classify all triads in the network: igraph
igraph::triad.census(gotbook.ig)

#Classify all triads in the network: statnet
sna::triad.census(gotbook.stat)
```

As with `dyad.census()` in the previous section, each library provides its own `triad.census()` function, and as such need the appropriate namespace specified if both are loaded. statnet's output includes the names of the triads, in addition to the counts.

Note: There are [16 possible classes](https://www.researchgate.net/figure/The-sixteen-isomorphism-classes-of-triad-census-for-directed-networks-D-stands-for_fig2_328894419) of directed triads (see Wasserman and Faust, fig. 6.7), only four of which are possible in undirected graphs. igraph considers this context undefined, and will output a warning message, as seen above, if `triad.census` is called on an undirected graph. Both igraph and statnet will only output the results for these four undirected permutations in undirected graphs, as the others cannot be determined.

Additionally, the R documentation for `statnet::triad_census()` contains (as of July 2022) the worryingly vague warning that 

> Valued data may cause strange behavior with this routine.

This suggests that weighted networks may result in unexpected results in a triad census in statnet. One way to sanity-check this is to compare the total count of triads against the number of possible triads in a network of size $n$ (having $n$ vertices). Recall that there are 

$$\binom{n}{k}$$

possible groups of size $k$ in a network of size $n$. So if the possible combinations of $k = 3$ vertices from the $n = 298$ total network vertices is the same as the sum of the triad types in the census, then the weighting is not affecting the output.

$$\binom{n}{k} = \binom{298}{3} = \frac{298 \times 297 \times 296}{3 \times 2 \times 1}$$

```{r triad-binary, echo=TRUE}
#Calculate possible triads in network
(298 * 297 * 296) / (3 * 2 * 1)

#total number of all triad types returned by triad census
sum(sna::triad.census(gotbook.stat))
```

In this case, the two calculations match, so we can be confident in the results. If they were different, the next step would be to remove the weights and repeat the process on the unweighted version of the network.

### Exercise: Triad Census

Now, using commands from either statnet or igraph, conduct a triad census of the `climate` network. Be sure to use the correct network data format (and name) for the command you choose to use.

```{r triad_climate, exercise=TRUE}
#Classify all triads in the network

```

```{r triad_climate-solution}
#Classify all triads in the network: igraph
igraph::triad.census(climate.ig)

#Classify all triads in the network: statnet
sna::triad.census(climate.stat)
```

## Transitivity or Global Clustering

While a triad census is interesting, it is highly descriptive and we may want a shorter reference point to understand the level of transitivity in the network. **Transitivity** is a network-level statistic that captures the proportion of connected triads in the network that are complete (all three possible links present), ranging from 0 (meaning none of the connected triads are transitive) to 1 (all connected triads are transitive). Note that there is a [weighted network method (Barrat)](https://toreopsahl.com/tnet/weighted-networks/clustering/) available in `igraph::transivity()`, but not explored here. 

```{r, echo=TRUE}
#get network transitivity: igraph
transitivity(gotbook.ig)
```

### Exercise: Transitivity

Now, compute the **global** transitivity of the climate network. Comparing results from the triad census to the overall transitivity measure, what have we learned about the network structure?

```{r trans, exercise=TRUE}
#get network transitivity

```

```{r trans-solution}
#get global network transitivity: igraph
transitivity(climate.ig)
```

Note that there is a also a command for transitivity in statnet/sna, `gtrans()` but that it is calculating a transitivity score based on an understanding of network structure rooted in hierarchy - it requires directed networks. Try calculating the hierarchical transitivity measure for the climate network using the `gtrans` command in statnet. 

```{r gtrans, exercise=TRUE}
#get hierarchical network transitivity:statnet

```

```{r gtrans-solution}
#get hierarchical network transitivity: statnet
gtrans(climate.stat, diag = T)
```

Note, there are also methods for weighted networks available in `sna::gtrans()`, rank and correlation, along with the relevant references.

## Local Transivity or Clustering

As we will explore more in future weeks, actors may be affected by the level of transitivity within their local ego networks. **Local transitivity**, more commonly known as the **local clustering coefficient**, is a technical description of how many of your friends are also friends with each other - it is just the density of an ego network. Let's compare the transitivity of the local ego networks of Petyr Baelish, Jon Snow, and Daenerys Targaryen in the Game of Thrones book network.

```{r, echo=TRUE}
#Retrive a list of the three vertices we are interested in
V(gotbook.ig)[c("Petyr Baelish", "Jon Snow", "Daenerys Targaryen")]

#check ego network transitivity: igraph
transitivity(gotbook.ig, type = "local", 
             vids = V(gotbook.ig)[c("Petyr Baelish", "Jon Snow", "Daenerys Targaryen")]) 

```

Note that the average of local transitivity for all vertices in a graph is often described as the **average clustering coefficient**, while the **global clustering coefficient** is equivalent to the transitivity of the entire network. The transitivity ratio places more emphasis on high-degree nodes, while the average clustering coefficient of a network places more emphasis on low-degree nodes. Compare the average clustering coefficient of the gotbook data to the transitivity ratio or global clustering coefficient.

```{r, echo=TRUE}
#get global clustering cofficient: igraph
transitivity(gotbook.ig, type = "global")
##get average local clustering coefficient: igraph
transitivity(gotbook.ig, type = "average")

```

### Exercise: Local Clustering

Compute the **local** clustering coefficient of three other vertices in the Game of Thrones book network: Tyrion Lannister, Cersei Lannister and Arya Stark. What do the differences in values between these two vertices, and the example vertices, say about these members of the network?

```{r localtrans, exercise=TRUE}
#check ego network transitivity: igraph

```

```{r localtrans-solution}
#check ego network transitivity: igraph
transitivity(gotbook.ig, type = "local", 
             vids = V(gotbook.ig)[c("Tyrion Lannister", "Cersei Lannister", "Arya Stark")]) 
```

Now, try comparing the global clustering coefficient to the average local clustering coefficient for the climate network.

```{r cluster, exercise=TRUE}
#get global clustering cofficient: igraph

#get average local clustering coefficient: igraph

```

```{r cluster-solution}
#get global clustering cofficient: igraph
transitivity(climate.ig, type="global")
#get average local clustering coefficient: igraph
transitivity(climate.ig, type="average")

```

## ADVANCED: Path Length and Geodesic

We often are interested in the shortest path, or distance, between two nodes. The igraph `distances` command calculates the distances between two or more nodes. Note that `distances` roughly corresponds to the concept of '6 degrees of separation' when the network ties are binary or unweighted, but the interpretation is quite different when edges are weighted. *Note that we must explicitly tell igraph to ignore edge weights* when calculating simple distances in weighted networks.

```{r, echo=TRUE}
#Calculate distances between two nodes
distances(gotbook.ig, "Petyr Baelish", "Robb Stark")

#Calculate distance using unweighted edges
distances(gotbook.ig, "Petyr Baelish", "Robb Stark", weights = NA)
```

We can also find all of the shortest paths between two characters in this network. These list the starting node (Bronn) and the ending node (Varys) with all nodes in between for each path. *The option `weights=NA` means that any available edge weights are ignored.*

```{r, echo=TRUE}
# list shortest paths between 2 nodes
all_shortest_paths(gotbook.ig, "Bronn", "Varys", weights = NA)$res
```

The above example shows that there are no intermediary nodes between Bronn and Varys. The following example shows three identical-length paths, each with one intermediary node, between Arys Oakheart and Robb Stark:

```{r, echo=TRUE}
all_shortest_paths(gotbook.ig, "Arys Oakheart", "Robb Stark", weights = NA)$res
```

While this command is useful for isolating the distances between specific nodes, it is often more useful to use the concept of shortest path to describe the overall network structure. The following calculates the average path length in the network as a whole. 

```{r, echo=TRUE}
#find average shortest path for network
average.path.length(gotbook.ig, directed = FALSE)
```

Compare the results of the two previous commands. What can you learn?

## ADVANCED: Component Structure and Membership

First, let's look at the network component structure. This gives a list with: membership by node, size of each component, and the total number of components.

```{r, echo=TRUE}
#What element are returned by components
names(igraph::components(gotbook.ig))
igraph::components(gotbook.ig)$no #Number of components
igraph::components(gotbook.ig)$csize #Size of each component
```

We can see from this that there is one large component with 195 members, and 103 components with only one member. We can easily find out the names of all the isolates in this component structure. Based on the boundaries put on nodes and ties in this dataset, what sense can we make of isolates in this dataset?

```{r, echo=TRUE}
#retrieve the index of isolate nodes
isolates(gotbook.stat)
#subset vertex.names attribute to get names of isolates
as.vector(gotbook.stat%v%'vertex.names')[c(isolates(gotbook.stat))]
```

### Exercise: Component Structure and Membership

Now, let's do the same inspections for the climate data.
Retrive the size of each component and the total number of components.

```{r structure, exercise=TRUE}
#get number of components

#get size of each component

```

```{r structure-solution}
igraph::components(climate.ig)$no #Number of components
igraph::components(climate.ig)$csize #Size of each component
```

And now, analyze the isolates, if any.

```{r isolates, exercise=TRUE}
#retrieve the index of isolate nodes
#subset vertex.names attribute to get names of isolates
```

```{r isolates-solution}
#retrieve the index of isolate nodes
isolates(climate.stat)
```