---
title: "Tutorial: Week 2"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
library(igraph)
library(statnet)
load("got_like_dislike.rda")
tutorial_options(exercise.timelimit = 10)
```


##**Introduction**
This script will introduce commands to do the following

* Conducting a dyad census of a network 
* Investigating node adjacency and distances
* Investigating walks and paths in the network 
* Conducting a triad census of a network
* Investigating component structure and membership

##**Step 1: Conducting a dyad census of a network**

We will make use of igraph methods and our igraph network object. Let's look at some of the edges
```{r}
    E(network_igraph)
```
    
How many edges are there?
```{r}
    ecount(network_igraph) 
```
    
What edge attributes do we have in the network? Edge weights? Types? Colors?

```{r}  
edge_attr(network_igraph) 
```


Let's try a full dyad census (command below). This gives us the number of: (a) Reciprocal (mutual), (b) Asymmetric (non-mutual), and (c) Absent relationships. If using an undirected graph, we will get a warning message using this, because it reports based on directed graphs. In an undirected graph, all edges are mutual and none are asymmetrical - but this distinction exists when directed

```{r, echo=TRUE}
igraph::dyad.census(network_igraph)
igraph::dyad.census(network_igraph)$mut
```

The number of reciprocal relationships (all of the edges, if the network is undirected)
```{r}
    igraph::dyad.census(network_igraph)$asym #The number of asymmetric relationships (none, if edges are not directed)
    igraph::dyad.census(network_igraph)$null #This is the number of pairs which could occur in in this network if directed
```

The `graph.density()` command gives the proportion of possible ties present in a network, without the math

```{r}
    graph.density(network_igraph,loops=F)
    1-graph.density(network_igraph,loops=F)
```
    
##**Step 2: Investigating node adjacency and distances**

**Adjacency**

```{r}
    interesting_node <-"Petyr Baelish" #Let's pick out an interesting character in the network
    neighbors(network_igraph,V(network_igraph)[interesting_node]) #Who are they connected to?

    adjacent_vertices(network_igraph,V(network_igraph)[interesting_node]) #This does the same thing as neighbors... 
    two_interesting_nodes<-c("Petyr Baelish","Varys") #but can handle multiple nodes...
    adjacent_vertices(network_igraph,V(network_igraph)[two_interesting_nodes]) #outputting a list for each.
```

**Distances**
There are several ways to measure the distance from one node to another, or all. The command `distances()` gives the length of all shortest paths from one node to one, some, or the rest of them
```{r}
    distances(network_igraph,"Petyr Baelish") 
```

  This command gives us a named numeric vector of distances from all nodes to one
```{r}
    distancesIn<-distances(network_igraph, "Petyr Baelish",mode="in") #We can specify this only to use paths pointing in...
    distancesOut<-distances(network_igraph,"Petyr Baelish",mode="out") #...or outward facing paths
    distancesIn==distancesOut #Notice that these are the same in an undirected graph - but they differ with a directed one

    distances(network_igraph,c("Petyr Baelish","Varys")) 
```

We can also specify more than one node in the network. When we have multiple nodes to get distances to/from, we get a numeric named matrix. Furthermore, we can calculate all distances across all nodes

```{r}
    dist_matrix<-shortest.paths(network_igraph,v=V(network_igraph),to=V(network_igraph))
    View(dist_matrix) 
    dist_matrix["Petyr Baelish","Varys"] #We can call on the row and column with the nodes we want to connect!
    dist_matrix["Petyr Baelish","Robb Stark"]
```

We can calculate the average path length in the network as a whole
```{r}
    average.path.length(network_igraph,directed=F)
```

Finally, notice that these results are different when the network is unweighted, and only binary

```{r}
    network_binary<-network_igraph #Create a binary version of the existing network
    E(network_binary)$weight<-1 #Assign all edge weights to be 1, not uniquely weighted


    dist_matrix_weights<-distances(network_igraph,v=V(network_igraph),to=V(network_igraph)) #Distances with weights
    dist_matrix_binary<-distances(network_binary,v=V(network_binary),to=V(network_binary)) #Without
    View(dist_matrix_weights) #Look at both matrices to spot the differences in path lengths
    View(dist_matrix_binary) #This is because algorithms for path lengths account for edge weight in short path calculations
```

**ADVANCED:**

```{r}
sum(distances(network_igraph,interesting_node) > average.path.length(network_igraph,directed=F)) #How many of Baelish's ties are longer than average?
sum(distances(network_igraph,interesting_node) < average.path.length(network_igraph,directed=F)) #How many are shorter than average?
```

We can also use the command `shortest.paths()` for the same result as `distances()`

```{r}
     shortest.paths(network_igraph,interesting_node) #and with the same in/out/all specifiers
     shortest.paths(network_igraph,interesting_node,mode = "in")
     shortest.paths(network_igraph,interesting_node,mode = "out")

     sum(dist_matrix_weights==dist_matrix_binary) #How many of the distance values are the same?
     sum(dist_matrix_weights!=dist_matrix_binary) #How many of the distance values are different?
```
     
##Step 3: Investigating walks and paths in the network
Let's meander around the network on what's called a 'walk', or random path of specified length

```{r}
    random_walk(network_igraph,start = interesting_node,steps=10) #This returns a list of all characters on this walk
    random10<-random_walk(network_igraph,start = interesting_node,steps=10)
    random10 #This lists all of the characters, even duplicates
    unique(random10) #Notice, some nodes may show up more than once on our random walk - fewer unique nodes
    random100<-random_walk(network_igraph,start=interesting_node,steps=100)
    unique(random100) #On longer random walks, we hit more nodes
```

We can also find all of the shortest paths between two characters in this network

```{r}
    node1<-"Bronn"
    node2<-"Varys"
    all_shortest_paths(network_igraph,node1,node2) #Notice, the output is a list of all possible paths
```

These list the starting node (Bronn) and the ending node (Varys) with all nodes inbetween for each path

**ADVANCED:**
Let's try creating a subgraph from a random walk - only include nodes on that walk
```{r}
     steps<-50 #We can change this value as we want
     subgraph_nodes<-random_walk(network_igraph,start=interesting_node,steps=steps) #This gives us nodes for the subgraph
     random_subgraph<-induced.subgraph(network_igraph,subgraph_nodes) #This subsets a network to specified nodes
```

   How does this network differ from the larger one?

```{r}
     graph.density(network_igraph)
     graph.density(random_subgraph)
 
     average.path.length(network_igraph)
     average.path.length(random_subgraph)
```
     
     How does this change with a different number of steps, or different start node? Give it a shot!
     
[insert practice exercise]

##Step 4: Conducting a triad census of a network
We can also get a proportion of triads in the network which are transitive
```{r}
transitivity(network_igraph) #Bound between 0 and 1, with 0 meaning none and 1 meaning all possible
```

We can even get local, actor level transitivity measures, which is simply the ratio of the triandles connected to a node to the triangles centered on it

```{r}
    transitivity(network_igraph,type="local",vids=V(network_igraph)[interesting_node]) #How 'transitive' is Petyr.Baelish?
    transitivity(network_igraph,type="local",vids=V(network_igraph)[neighbors(network_igraph,interesting_node)]) #What about his neighbors?
```

**ADVANCED:**

Let's compare the local transitivity of eacch node with the global average
1. Assign a node value for local transitivity

```{r}
     V(network_igraph)$local_transitivity<-transitivity(network_igraph,type="local",vids=V(network_igraph))
```

2. Assign a node value for local transitivity normalized by global

```{r}
     V(network_igraph)$relative_local_transitivity<-transitivity(network_igraph,type="local",vids=V(network_igraph))/transitivity(network_igraph)
```

3. Build a dataframe with these data

```{r}
     transitivityDF<-data.frame(name=V(network_igraph)$name,
                               local_transitivity=V(network_igraph)$local_transitivity,
                               relative_local_transitivity=V(network_igraph)$relative_local_transitivity)
     View(transitivityDF)
     transitivity(network_igraph)
     mean(transitivityDF$local_transitivity,na.rm=T)
     mean(transitivityDF$relative_local_transitivity,na.rm=T)
```
     
##Step 5: Investigating component structure and membership
Let's first make two different networks, one with only 'likes' and one with only 'dislikes'

```{r}
    network_like<-igraph::delete.edges(network_igraph,edges = E(network_igraph)[E(network_igraph)$type=="dislike"])
    network_dislike<-igraph::delete.edges(network_igraph,edges = E(network_igraph)[E(network_igraph)$type=="like"])
```

First, let's look at the 'like' network component structure:

```{r}
    components(network_like) #This gives a list with: membership by node, size of each component, and the total number of components
    components(network_like,mode = "weak") #Strong has different threshold for ties than weak, 
    components(network_like,mode = "strong") #so it usually offers more components than the weak mode does

    components(network_like,mode = "strong")$membership #Membership by node
    components(network_like,mode = "strong")$csize #Size of each component
    components(network_like,mode = "strong")$no #Number of components
```
  
  Which one is the major component in the 'like' network?
```{r}
    components(network_like,mode="strong")$csize #It's the first one, which has 43 of the 46 nodes
    components(network_like,mode="strong")$csize/vcount(network_like) #This gives the proportion of nodes in each component
```

Who's an isolate in this component structure? Let's look at the component with one member

```{r}
    components(network_like,mode = "strong")$csize #That's the fourth one
    node_membership<-components(network_like,mode = "strong")$membership
    node_membership[which(node_membership==4)] #Viserys, a Targaryen! No wonder...
```

 [make into exercise; repeat for dislike] Let's quickly repeat this with the 'dislike' network#
    components(network_dislike,mode = "weak")
    components(network_dislike,mode = "strong")
    components(network_dislike,mode = "strong")$csize #Which is largest? Which has isolates?
    node_membership2<-components(network_dislike,mode = "strong")$membership
    node_membership2[which(node_membership2 %in% c(2:8))] #Nodes in  one-node components

**ADVANCED:**

Finally, let's create and plot subgraphs of these networks with only the major components. We can assign component membership as a node attribute in each network
```{r}
     V(network_like)$comp<-node_membership
     V(network_dislike)$comp<-node_membership2
```
 
We can then create subgraphs in new objects by deleting non-major component nodes from the originals
```{r}
     network_like_major<-igraph::delete.vertices(network_like,V(network_like)[V(network_like)$comp!=4])
     network_dislike_major<-igraph::delete.vertices(network_dislike,V(network_dislike)[V(network_dislike)$comp!=1])
```
 
First, let's assign a color to each faction to see them in the networks:

```{r}
    house_colors<-data.frame(faction=unique(V(network_igraph)$faction),color=NA) #Data frame with all unique houses
     house_colors$color<-rainbow(nrow(house_colors)) #Assign a unique color to each faction, or the lack of a house
     V(network_like)$color<-house_colors$color[match(V(network_like)$faction,house_colors$faction)] #Create node attributes
     V(network_dislike)$color<-house_colors$color[match(V(network_dislike)$faction,house_colors$faction)] #in both networks
```
 
Let's plot them before and after to see how they look:
Like network, total vs major component
```{r}
   plot(network_like, #First part of plot function is network to plot,
         vertex.label=NA, #Then the labels, or NA if you want none
         vertex.color=V(network_like)$color, #The node-level color attribute we assigned
         edge.arrow.size=0.5) #Size of the arrows (they often get in the way)
     plot(network_like_major,
         vertex.label=NA,
         vertex.color=V(network_dislike)$color,
         edge.arrow.size=0.5)
```
[make dislike plot into exercise:]
Dislike network, total vs major component
     plot(network_dislike,
         vertex.label=NA,
         vertex.color=V(network_dislike)$color,
         edge.arrow.size=0.5)
     plot(network_dislike_major,
         vertex.label=NA,
         vertex.color=V(network_dislike)$color,
         edge.arrow.size=0.5)


