---
title: "Tutorial: Network Degree, Week 3"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(igraph)
library(statnet)
library(tidyverse)
load("Week3.rdata")
tutorial_options(exercise.timelimit = 10)
```

## Introduction
This tutorial will demonstrate use of the R commands used to measure aspects of network data introduced in Week 3's lecture:

* Network density
* Vertex degree
* Vertex degree in directed networks
* Vertex degree summary statistics
* Vertex degree distribution
* Identifying outlier nodes
* Network centralization

We will be using two datasets for this exercise. The first, Swiss Climate Influence network, was used in the Week 2 tutorial. The climate data includes 'loops', or ties from a node to itself because some states have nominated themselves. This will affect some of our calculations, but we will get to that later. The second is taken from the trade data included with the Correlates of War data, and the original data is based on annual trade flows between nation-states in 2007. From the description: "The dyadic trade dataset describes import and export data in current U.S. dollars for pairs of sovereign states."

- *climate.ig* - igraph network object
- *climate.stat* - statnet network object
- *trade2007.ig* - igraph network object
- *trade2007.stat* - statnet network object
- *trade100.stat* - (0/1) tie when nation imports >100million

## Basic Network Description

Inspect the basic network descriptors for the `trade2007` network. For this exercise, refer to tutorials from Weeks 1 and 2. This will help you get familiar with the dataset you are using.

1. **Describe the Network:** Find network size, type (un/directed, un/weighted, bipartite) and available attributes of vertices and edges.

```{r describe, exercise=TRUE}
#Find network size and type
```

```{r describe-solution}
#Find network size (vertex and edge count): igraph
vcount(trade2007.ig)
ecount(trade2007.ig)

#Find network size (vertex and edge count): statnet
print(trade2007.stat)
# Or:
network.size(trade2007.stat)
network.edgecount(trade2007.stat)
```

2. **Dyad Census:** Now, use the tutorial from Week 2 to inspect the dyads in the network. 

```{r dyad, exercise=TRUE}
#Classify all dyads in the network

```

```{r dyad-solution}
#Classify all dyads in the network: igraph
igraph::dyad.census(trade2007.ig)

#Classify all dyads in the network: statnet
sna::dyad.census(trade2007.stat)
```

3. **Triad Census:** Are there any interesting patterns for triads in the network?

```{r triad_climate, exercise=TRUE}
#Classify all triads in the network

```

```{r triad_climate-solution}
#Classify all triads in the network: igraph
igraph::triad.census(trade2007.ig)

#Classify all triads in the network: statnet
sna::triad.census(trade2007.stat)
```


4. **Clustering and Transitivity**: Calculate the global and average clustering coefficient, as well as the network transitivity.

```{r cluster, exercise=TRUE, exercise.lines=6}
#get global clustering coefficient: igraph

#get average local clustering coefficient: igraph

#get network transitivity: statnet

```

```{r cluster-solution}
#get global clustering coefficient: igraph
transitivity(trade2007.ig, type="global")
#get average local clustering coefficient: igraph
transitivity(trade2007.ig, type="average")
#get network transitivity: statnet
gtrans(trade2007.stat)
```

5. **Component Structure and Isolates**: Inspect the component structure of the network. Are there any isolates?

```{r structure, exercise=TRUE, exercise.lines=4}
#get number of components

#get size of each component

```

```{r structure-solution}
igraph::components(trade2007.ig)$no #Number of components
igraph::components(trade2007.ig)$csize #Size of each component
```

## Density

The *density* of a network is defined as the ratio of the number of edges it has, to the maximum number it *could* have if all nodes were connected. The question of whether loops, or self-connected nodes, should be considered in density calculations is an academic one without a clear general answer; as a default, it is probably best not to consider loops in density calculations unless it makes sense in your application. Since every node could potentially connect to itself as well as the other nodes in the network, factoring this possibility in will affect the output.

(32 of 34 actors in the climate network do connect to themselves, so we will have to contend with their presence in our calculations in this example.)

Note: edge *weights* are not part of the calculation, and are not considered in either igraph or statnet. 
igraph provides the `graph.density()` function, a wrapper around (and synonym for) `edge_density`. Its optional argument `loops`, which defaults to `FALSE`, tells it whether or not to make the calculation considering loops. Note that it does *not* cause existing loops not to be counted; it presumes that you know whether there are loops in the network or not. If you do not want to include loops in the calculation, it is best to first remove them via igraph's `simplify()` function, which also removes multiple edges.

```{r, echo=TRUE}
#get network density: igraph
graph.density(climate.ig)

# Get network density with igraph, omitting loops
# (Be sure to use the igraph:: namespace with simplify if also using 
# purrr/tidyverse, to avoid collisions with purrr::simplify)
graph.density(igraph::simplify(climate.ig))
```

Statnet provides two different functions: `network.density()` and `gden()`. `gden()` has an optional `diag` argument, defaulting to `FALSE` as with igraph's `loops` argument in `graph.density()` above, which describes whether to count loops (the diagonal of a network matrix is where loops are found, nodes that connect to themselves). Note that unlike igraph's `graph.density()`, it does not require loops to be removed from the network prior to the calculation. (It also has a `mode()` argument that defaults to `digraph`, for directed network; set it to `graph` if the network is undirected.)

```{r echo=TRUE}
# get network density: statnet
# the default in gden() is to omit loops
gden(climate.stat, diag = FALSE, mode = "digraph")
```

`network.density()` assumes and counts loops in the network, and is the equivalent of igraph's `graph.density()` with `loops = TRUE` for networks with loops.

```{r echo=TRUE}
# compare igraph's graph.density(x, loops = TRUE) with statnet's 
# network.density with default settings
igraph::graph.density(climate.ig, loops = TRUE)
network::network.density(climate.stat)
```

*Note: `gden()` and `network.density()` deal with missing edges differently. The former omits missing edges from numerator but not denominator, the latter omits from both.*

### Exercise: Network Density

Calculate network density for the trade2007 network. Given the attributes of this network, does it matter which package you use for this calculation? Based on the loops explanation for the climate output, could this account for the difference? What else could be going on? You may want to go back and look carefully at both versions of the trade2007 network. Does the answer change if you use the binary network where ties exist only if trade is more than 100 million?

```{r density, exercise=TRUE}
#get network density: igraph or statnet

```

```{r density-solution}
#get network density: igraph
graph.density(trade2007.ig)

#get network density: statnet
network.density(trade2007.stat)

#get network density: statnet
network.density(trade100.stat)
```

## Vertex Degree

Perhaps the simplest measure of node centrality or popularity is degree: a count of the relationships that a node is involved in, or the count of the edges that are incident to a vertex. This is similar to counting how many other nodes a single node is connecting to, although there is the possibility of self-connection (loops, as previously discussed), and multiple connections to specific single nodes, that can complicate the picture somewhat. It is more properly described as the number of edges a node connects to, whether those edges are to other nodes, or to themselves.

```{r, echo=TRUE}
#Calculate average network degree: igraph
igraph::degree(climate.ig)

#Calculate average network degree: statnet
sna::degree(climate.stat)
```

Two differences between the two commands are clearly visible. First, `sna::degree` ignores loops or self-nominations, while `igraph::degree` includes loops. Including the option `loops = FALSE` in the `igraph::degree` command will make the two commands equivalent for this data set. Note that setting `diag = TRUE` in `sna::degree` does not guarantee equivalence, as statnet only single counts the loops in a directed network, while igraph double-counts the loops.

Second, the igraph command also provides node name along with node degree, which can be helpful if you would like to know the degree of a particular node. As you can imagine, however, this list can be quite difficult to interpret in datasets with many nodes, or when nodes have long names. Therefore, we rarely use the degree command to list the degree of all nodes, but instead use it to:

* compute network summary statistics, such as average network degree
* graph the network degree distribution, or
* identify particular nodes, such as the *n* nodes with the most contacts.

## Review: Create a Dataframe

Regardless of the way(s) in which we plan to use network degree measures, it will be useful to store vertex attributes such as degree in a table of node attributes. This table, or dataframe, of node measures can even be used in traditional econometric models that include network structure variables. Here is an example of the basic R command to create a dataframe with named variables `names` and `degree`, and row names set as the entries in the `names` variable. 

```{r, echo=TRUE}
# create a dataset of vertex names and degree: igraph
climate.nodes <- data.frame(name = V(climate.ig)$name, 
                            degree = igraph::degree(climate.ig))
climate.nodes
```

Here are the same commands in igraph, but not evaluated to avoid repetition.

```{r, echo=TRUE, eval=FALSE}
# create a dataset of vertex names and degree: statnet
climate.nodes <- data.frame(name = climate.stat %v% "vertex.names", 
                            degree = sna::degree(climate.stat))
```

For those who need a quick breakdown of what these commands do:

* `climate.nodes` is a dataframe that we are creating within R. 
* `names` and `degree` are two variables (columns) in `climate.nodes`
* the `names` variable is made up of the vertex names from `climate.ig` or `climate.statnet`
* the `degree` variable is calculated from the climate dataset.

### Exercise: Create a dataframe

You can do the same thing with the `trade2007` dataset. The goal is to create a dataframe of names and degrees, so using the code above, simply replace the climate elements with `trade2007`.

```{r df, exercise=TRUE}
# create a dataset of vertex names and degree: igraph or statnet

```

```{r df-solution}
# create a dataset of vertex names and degree: igraph
trade2007.nodes <- data.frame(name = V(trade2007.ig)$name, 
                              degree = igraph::degree(trade2007.ig))

trade2007.nodes # output the dataframe

# create a dataset of vertex names and degree: statnet
trade2007.nodes <- data.frame(name = trade2007.stat %v% "vertex.names", 
                              degree=sna::degree(trade2007.stat))

trade2007.nodes # output the dataframe

```


## Degree in Directed Networks

In a directed network, it is possible to calculate both *indegree*, or the total number of nodes that send a tie to the focal node, and *outdegree*, or the total number of nodes to which the focal node sends a tie. R allows you to specify whether you want to calculate in-degree, out-degree, or total degree when working with a directed network. By default, the function `degree` produces a count of total degree.

```{r, echo=TRUE}
#calculate in-degree: statnet
sna::degree(climate.stat, cmode = "indegree")
#calculate out-degree: statnet
sna::degree(climate.stat, cmode = "outdegree")
```

Here are the same commands in igraph.

```{r, echo=TRUE}
#calculate in-degree: igraph
igraph::degree(climate.ig, mode = "in", loops = FALSE)
#calculate out-degree: igraph
igraph::degree(climate.ig, mode = "out", loops = FALSE)
```

To make use of these calculations, you'll want to attach them to the dataframe you created in the last exercise. To do this, we can use the `dplyr::mutate()` function to add columns onto the `climate.nodes` object created before. 

```{r, echo=TRUE}
#statnet version:
climate.nodes <- climate.nodes %>%
    mutate(indegree = sna::degree(climate.stat, cmode = "indegree"),
           outdegree = sna::degree(climate.stat, cmode = "outdegree")
           )

climate.nodes
```

Here are the same commands in igraph, but not evaluated to avoid repetition.

```{r, echo=TRUE, eval=FALSE}
#igraph version:
climate.nodes <- climate.nodes %>%
    mutate(indegree = igraph::degree(climate.ig, mode = "in", loops = FALSE),
           outdegree = igraph::degree(climate.ig, mode = "out", loops = FALSE))
```

### Exercise

Now, do the same for the trade2007 network. Create a dataframe that includes all three measures of degree: total degree, indegree, and outdegree. To do this, you'll simply need to add the codes of `totdegree`, `indegree` and `outdegree` to replace the initial `degree` column. You don't need to evaluate the degrees before attaching them to the dataframe. 

```{r degree2, exercise=TRUE}
#attach the in, out, and total degree values to the dataframe

```

```{r degree2-solution}
#igraph version:

trade2007.nodes <- data.frame(name = V(trade2007.ig)$name,
    totdegree = igraph::degree(trade2007.ig, loops = FALSE),
    indegree = igraph::degree(trade2007.ig, mode = "in", loops = FALSE),
    outdegree = igraph::degree(trade2007.ig, mode = "out", loops = FALSE))

trade2007.nodes
#statnet version:

trade2007.nodes <- data.frame(name = trade2007.stat %v% "vertex.names",
    totdegree = sna::degree(trade2007.stat),
    indegree = sna::degree(trade2007.stat, cmode = "indegree"),
    outdegree = sna::degree(trade2007.stat, cmode = "outdegree"))

trade2007.nodes
```

## Network Degree Summary Statistics

What does the distribution of these scores look like? We can use the basic `summary()` function to find the range, mean, median and standard deviation of a vector.

```{r, echo=TRUE}
#create a vector with the numbers 1 to 10
v <- 1:10
#summary statistics of x
summary(v)
```

Now, we can use the same function to get basic summaries for all the columns in a dataframe. In this case, we will get summary statistics for the climate network node attributes.

```{r, echo=TRUE}
#get summary statistics for node attributes
summary(climate.nodes)
```

### Exercise: Summary Statistics

Try getting summary statistics for node degree in the 2007 trade network.

```{r summ, exercise=TRUE}
#get summary statistics for trade network node attributes

```

```{r summ-solution}
summary(trade2007.nodes)

```

### ADVANCED Exercise

Try getting summary statistics for node degree for the binary network degree, creating the appropriate dataframe in this exercise and getting the summary. Unlike other exercises, we do not provide the correct dataframe for you to access!

```{r summadv, exercise=TRUE}
#get summary statistics for binary trade100 network node attributes

```

```{r summadv-solution}
summary(data.frame(name = trade100.stat %v% "vertex.names",
    totdegree = sna::degree(trade100.stat),
    indegree = sna::degree(trade100.stat, cmode = "indegree"),
    outdegree = sna::degree(trade100.stat, cmode = "outdegree")))

```

## ADVANCED: Summarise by Group

There are times when we may want to know about sub-groups of nodes, not the entire network. For example, maybe we are interested in average network characteristics of characters in the game of thrones grouped by family or age.  Here is a basic example of creating summary statistics for groups.

```{r, echo=TRUE}
#ADVANCED: summarize by type
df <- data.frame(scores = c(55, 23, 48, 3, 112, 14, 25), 
                 type = c("A", "B", "A", "B", "B", "A", "B"))

df %>%
  group_by(type) %>%
  summarise_all(funs(mean, n()))
```

### Exercise

Try attaching the `type5` vertex attribute to the climate node dataframe. Can you `summarise` the data after grouping by organization `type`?

```{r advanced, exercise=TRUE}
```

```{r advanced-solution, results="hide", message=FALSE}
#attach the additional column 
climate.nodes$type5 <- V(climate.ig)$type5

climate.nodes %>%
  group_by(type5) %>%
  select(indegree, outdegree) %>%
  summarise_all(., funs(mean, n()))
```

## Network Degree Distribution

The command `hist` is a basic R command that draws a histogram of vector of data. We can set options such as `main` for the main title of the histogram and `xlab` for the label of the x-axis. Here is a simple example.

Let's examine the histograms for the distributions of node in- and out-degrees for the climate network. What do you make of the differences in the two histograms?

```{r, echo=TRUE}
#create a histogram of Climate Influence Indegree
hist(climate.nodes$indegree, 
     main = "Climate Influence: In-degree Distribution", 
     xlab = "Nominations Received")

hist(climate.nodes$outdegree, 
     main = "Climate Influence: Out-degree Distribution ", 
     xlab = "Nominations Made")
```

### Exercise

Now, try creating histograms of in and out degree for the 2007 trade network.

```{r hist, exercise=TRUE}

```

```{r hist-solution}
hist(trade2007.nodes$indegree, 
     main = "2007 Trade Flows: In-degree Distribution", 
     xlab = "Import Connections")

hist(trade2007.nodes$outdegree, 
     main = "2007 Trade Flows: Out-degree Distribution", 
     xlab = "Export Connections")
```

## Network Degree Centralization

We can also calculate a network-level *centralization* score based on actor degree scores. (igraph and statnet both provide a `centralization()` function, so it is recommended to prefix the function name with the appropriate library to avoid conflicts.) 

The statnet function takes the argument `degree` (without quotes, a reserved word), to describe the function to use for the node scores; the `cmode` argument is passed to this `degree` function. Its valid values in this context are "indegree", "outdegree", and "freeman", which is the combined total of indegree and outdegree. It also has a `diag` argument, defaulting to `FALSE`, which describes whether the diagonal (or loops) should be counted in the calculation.

What might be the interpretation of different centralization scores for in-degree and out-degree in the climate network?

```{r, echo=TRUE}
#get network centralization score: statnet
sna::centralization(climate.stat, degree, cmode = "indegree")
sna::centralization(climate.stat, degree, cmode = "outdegree")
```

The igraph `centr_degree()` function returns an object with several components; we are interested in the `centralization` component. The valid `mode` argument options are "in", "out", or "total". Note that the `loops` argument defaults to `TRUE`, and should be set to `FALSE` if it is not desired to include loops in the calculation, to match statnet's default above.

```{r, echo=TRUE}
#get network centralization score: igraph
centr_degree(climate.ig, loops = FALSE, mode = "in")$centralization
centr_degree(climate.ig, loops = FALSE, mode = "out")$centralization
```

### Exercise: Network degree centralization

Calculate network degree centralization scores for both in-degree and out-degree using the 2007 trade network. Compare to the degree centralization scores for the binary trade network with a threshold of $100 million for a tie.

```{r central, exercise=TRUE}
#get network degree centralization

```

```{r central-solution}
centralization(trade2007.stat, degree, cmode = "indegree")
centralization(trade2007.stat, degree, cmode = "outdegree")
centralization(trade100.stat, degree, cmode = "indegree")
centralization(trade100.stat, degree, cmode = "outdegree")
```

## Identifying Specific Nodes

There may be times when we are interested in identifying and describing nodes with specific values, such as the nodes with the most or the fewest number of network ties. In order to identify these nodes, we will need to sort vectors and print out a subset of the possible nodes.

### Basic commands: sorting

As a reminder, we can use the `arrange()` function to sort data objects in R. (Note that many packages provide implementations of `arrange()` appropriate to their data types; the one shown here is from `dplyr`, part of the tidyverse.)

```{r, echo=TRUE}
#sort a vector
df <- data.frame(scores = c(55, 23, 48, 3, 112, 14, 25, 48), 
                 type = c("A", "B", "A", "B", "B", "A", "B", "B"))
arrange(df, scores)
```

Using the `desc()` option in `arrange` puts the largest values at the top of the list rather than the bottom.

```{r, echo=TRUE}
#sort largest numbers first
arrange(df, desc(scores))
```

The `slice` family of functions offer many ways to arrange and subset data. For example, the functions `slice_min()` and `slice_max()` can be used to sort a dataframe by a specific field and return a subset. The example below returns the rows with the 3 highest scores:

```{r, echo=TRUE}
#only return 3 largest values
df %>% slice_max(scores, n = 3)
```

Note that the output above listed 4 results, even though we specified `n = 3` in the `slice_max()` function. Why? Because the lowest value of the top 3, 48, was shared by 2 nodes, so they were both listed. This means that these functions can return more rows that are requested, as you will see in the final exercise. To suppress this, add `with_ties = FALSE` to the command, but be aware that it is undefined which of the tied answers will be omitted:

```{r, echo=TRUE}
#only return 3 largest values, suppress ties
df %>% slice_max(scores, n = 3, with_ties = FALSE)
```


### Exercise: sorting
Try sorting a vector of vertex degree for the climate data, and list the 5 nodes with the highest indegree. Can you also list the five nodes with the lowest indegree? How about the highest outdegree?

```{r degree1, exercise=TRUE}
#sort nodes in the climate network

```

```{r degree1-solution, eval=FALSE}
#sort nodes in the climate network
climate.nodes %>% slice_max(indegree, n = 5)
climate.nodes %>% slice_min(indegree, n = 5)
climate.nodes %>% slice_max(outdegree, n = 5)
```

### Exercise 2
Do the same thing for the `trade2007` dataset. For this, simply replace the dataframe name from the previous exercise:

```{r degree3, exercise=TRUE}
#sort nodes in the trade2007 network

```

```{r degree3-solution}
#sort nodes in the trade2007 network
trade2007.nodes %>% slice_max(indegree, n = 5)
trade2007.nodes %>% slice_min(indegree, n = 5)
trade2007.nodes %>% slice_max(outdegree, n = 5)
```

### ADVANCED Exercise
Can you find the low and high degree nodes for the binary trade network? You will need to use some creativity in figuring out the best way to code this.

```{r degreeadv, exercise=TRUE}
#sort to find 5 highest and lowest degree nodes in trade100

```

```{r degreeadv-solution}
#sort to find 5 highest and lowest degree nodes in trade100
temp <- data.frame(name = trade100.stat %v% "vertex.names",
    indegree = sna::degree(trade100.stat, cmode = "indegree"),
    outdegree = sna::degree(trade100.stat, cmode = "outdegree"))

temp %>% slice_max(indegree, n = 5, with_ties = FALSE)
temp %>% slice_min(indegree, n = 5, with_ties = FALSE)
temp %>% slice_max(outdegree, n = 5, with_ties = FALSE)
temp %>% slice_min(outdegree, n = 5, with_ties = FALSE)
```

